---
title: "Why Not to Use SPDs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why Not to Use SPDs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev='png',
  fig.width=7, 
  fig.height=5.5 # ,
  # dev.args=list(antialias = "none")
)
```

```{r setup}
library(carbondate)
```

# Summed Probability Distributions
Currently, the most commonly-used approach to summarise calendar age information from multiple ^14^C determinations is to calculate the summed probability distribution (SPD). However, SPDs provide **neither reliable nor statistically valid** estimation of the summarised calendar age information. It is our view that they should not be used in any dates-as-data approach to provide a population proxy. For complete details, with comprehensive illustrative examples, on why SPDs should not be used for summarisation and how they can provide highly misleading inference, see [@heaton2025]).

## The Problems with using SPDs
When creating an SPD, the posterior calendar age density of each object is first calculated independently from the others. These individual densities are then summed/averaged to give an SPD estimate. The independence assumed in the initial calibration of each sample, is fundamentally contradictory to the subsequent summarisation and results in a calendar age summary that is both unreliable and inconsistent. 

Additionally, the SPD approach fundamentally does not model the samples in the calendar age domain. Consequently, it is also not able to deal with inversions in the calibration curve where there are multiple disjoint calendar periods which are consistent with the observed determinations; or with plateau periods. 

The SPD function is **ONLY** provided here as a comparison with the other routines. To calculate the SPD for a set of radiocarbon determinations (here we use the example dataset `armit` [@armit2014]) see the example below, where we also plot the results.

```{r find_spd, out.width="100%"}
spd <- FindSummedProbabilityDistribution(
  calendar_age_range_BP = c(1000, 4500), 
  rc_determinations = armit$c14_age, 
  rc_sigmas = armit$c14_sig, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE)
```

**Note:** The functions we provide to plot the rigorous calendar age summary provided by our Bayesian non-parametric DPMM alternative to SPDs  can also optionally plot the SPD - see the vignette [Non-parametric Calendar Age Summarisation](Non-parametric-summed-density.html) for details.

## Illustration of why not to use SPDs

### Fitting to a mixture of two normal distributions

The `two_normals` dataset contains 50 simulated ^14^C determinations. These were created by first drawing a set of 50 calendar ages from a mixture of two normal calendar age densities - one centred at 3500 cal yr BP (with a 1$\sigma$ standard deviation of 200 cal yrs); and another (more concentrated) centred at 5000 cal yr BP (with a 1$\sigma$ standard deviation of 100 cal yrs). Having simulated these calendar ages, a corresponding ^14^C determination for each sample were created using the IntCal20 curve [@reimer2020]. 

When we summarise the calendar age information provided by the 50 simulated ^14^C determinations, we would aim to reconstruct the underying mixture of two normals that were used to generate the data. However, when we calculate the SPD we obtain:

```{r two_normals_spd, echo = 2,  out.width="100%"}
oldpar <- par(no.readonly = TRUE)

two_normals_spd <- FindSummedProbabilityDistribution(
  calendar_age_range_BP=c(2500, 7000),
  rc_determinations= two_normals$c14_age,
  rc_sigmas = two_normals$c14_sig,
  calibration_curve=intcal20,
  plot_output = TRUE)

par(new = TRUE, 
    mgp = c(3, 0.7, 0),
    xaxs = "i",
    yaxs = "i",
    mar = c(5, 4.5, 4, 2) + 0.1,
    las = 1)
xlim <- rev(range(two_normals_spd$calendar_age_BP))
ylim <- c(0, 3 * max(two_normals_spd$probability))
plot(
    NULL,
    NULL,
    type = "n",
    ylim = ylim,
    xlim = xlim,
    axes = FALSE,
    xlab = NA,
    ylab = NA,
    xaxs = "i",
    yaxs = "i")
# Show true underlying calendar age density
weights_true <- c(0.45, 0.55)
cluster_means_true_calBP <- c(3500, 5000)
cluster_precisions_true <- 1 / c(200, 100)^2

# Find and plot true exact density
truedens <- function(t, w, truemean, trueprec) {
  dens <- 0
  for(i in 1:length(w)) {
    dens <- dens + w[i] * dnorm(t, mean = truemean[i], sd = 1/sqrt(trueprec[i]))
  }
  dens
}
curve(truedens(
  x,
  w = weights_true,
  truemean = cluster_means_true_calBP,
  trueprec = cluster_precisions_true),
      from = 2500, to = 7000, n = 401,
      lwd = 2,
      col = "red", add = TRUE)

# Reset plotting parameters
par(oldpar)
```

Here we have manually overlain the true (in this case, known) shared calendar age density in red. As we can see, the SPD captures does capture some broad features of that underlying calendar age distribution but does not reconstruct the truth well, and is hard to interpret. In particular, the SPD is highly variable, showing multiple peaks, due to the wiggliness of the calibration curve. The SPD peak shown around 5300 cal yr BP is entirely spurious, yet almost of the same magnitude as its peak around 3500 cal yr BP (which is a part of the genuine density). 

## An improvement using our library approaches

While this is jumping forward somewhat, to evidence that our methods provide better reconstructions, we run the same example using our Bayesian non-parametric summarisation approach (shown in purple as _Polya Urn_) and obtain:

```{r two_normals_compare, echo = FALSE,  out.width="100%"}
oldpar <- par(no.readonly = TRUE)

polya_urn_output <- PolyaUrnBivarDirichlet(
  rc_determinations = two_normals$c14_age,
  rc_sigmas = two_normals$c14_sig,
  calibration_curve=intcal20,
  n_iter = 1e4,
  show_progress = FALSE)

two_normals_DPMM <- PlotPredictiveCalendarAgeDensity(
  output_data = polya_urn_output,
  show_SPD = TRUE)

predictive_density <- two_normals_DPMM$predictive_density

par(new = TRUE, 
    mgp = c(3, 0.7, 0),
    xaxs = "i",
    yaxs = "i",
    mar = c(5, 4.5, 4, 2) + 0.1,
    las = 1)
xlim <- rev(range(predictive_density[[1]]$calendar_age_BP))
ylim <- c(0, 3 * max(predictive_density[[1]]$density_mean))
plot(
    NULL,
    NULL,
    type = "n",
    ylim = ylim,
    xlim = xlim,
    axes = FALSE,
    xlab = NA,
    ylab = NA,
    xaxs = "i",
    yaxs = "i")
# Show true underlying calendar age density
weights_true <- c(0.45, 0.55)
cluster_means_true_calBP <- c(3500, 5000)
cluster_precisions_true <- 1 / c(200, 100)^2

# Find and plot true exact density
truedens <- function(t, w, truemean, trueprec) {
  dens <- 0
  for(i in 1:length(w)) {
    dens <- dens + w[i] * dnorm(t, mean = truemean[i], sd = 1/sqrt(trueprec[i]))
  }
  dens
}
curve(truedens(
  x,
  w = weights_true,
  truemean = cluster_means_true_calBP,
  trueprec = cluster_precisions_true),
      from = 2500, to = 7000, n = 401,
      lwd = 2,
      col = "red", add = TRUE)

# Reset plotting parameters
par(oldpar)
```

# References
