---
title: "Calibrating Marine 14C Samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating Marine 14C Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev='png',
  fig.width=7, 
  fig.height=5.5 # ,
  # dev.args=list(antialias = "none")
)
```

```{r setup}
library(carbondate)
set.seed(15)
```

# Calibrating marine ^14^C determinations 

<!-- badges: start -->
<!-- badges: end -->

It is possible to calibrate and summarise marine $^{14}$C determinations within the library. Marine $^{14}$C determinations are slightly more complex to calibrate since they require a localised $\Delta R$ adjustment before calibration against the Marine20 calibration curve, see @heaton2023 for details on marine calibration and @heaton2020 for the Marine20 calibration curve.    

*Note:* You must use the specific $\Delta R$ adjustment appropriate for the particular marine calibration curve you are using for calibration (as the estimates of $\Delta R$ change each time the marine calibration curve is updated). The most recent estimates  $\Delta R_{20}$, calculated based on the Marine20 calibration curve, can be found at http://calib.org/marine/ 


## Calibrating a single marine ^14^C sample

As described in the [independent calibration vignette](Independent-calibration.html), we can calibration an individual marine $^{14}$C sample in `CalibrateSingleDetermination()` by specifying the inbuilt `marine20` calibration curve [@heaton2020], and providing an estimate for the localised $\Delta R$ (and its $1\sigma$-uncertainty by specifying `delta_r` and `delta_r_sig`: 

```{r single_calibration_Marine_example, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = marine20,
  delta_r = 150,
  delta_r_sig = 50, 
  plot_output = TRUE)
```

Here the radiocarbon age axis shows the analytical ^14^C determination in red, while the adjusted determination (after applying a $\Delta R = 150 \pm 50$ ^14^C yrs) is shown in green. Intuitively, we calibrate the green (adjusted) value against the Marine20 curve. 

**Note:** One can also use the `delta_r` and `delta_r_sig` arguments to specify a general offset to a calibration curve (i.e., these arguments can also be used when using the IntCal and SHCal curves if the samples is thought to be offset).    

## Non-Parametric Joint Density Estimation from a set of marine ^14^C samples
If we have a set of marine ^14^C samples which share a common $\Delta R$ then we can use the `PolyaUrnBivarDirichlet()` or `WalkerBivarDirichlet()` functions to obtain a non-parametric DPMM summary of the joint (predictive) calendar age distribution. To do this we specify the value of $\Delta R$ and uncertainty as additional arguments, and state that we wish to use the marine radiocarbon calibration curve. 

### Example - Two Normals Marine Data 
We consider the artificial `two_normals_marine` dataset which contains 50 simulated ^14^C samples. This dataset was created by drawing 50 calendar ages from a mixture of two normal densities - one centred at 3500 cal yr BP (with a 1$\sigma$ standard deviation of 200 cal yrs); and another (more concentrated) centred at 5000 cal yr BP (with a 1$\sigma$ standard deviation of 100 cal yrs). 

The corresponding 50 $^{14}$C determinations are then simulated based upon the Marine20 calibration curve [@heaton2020] with an assumed $\Delta R_{20}$ of $-50 \pm 30$ $^{14}$C yrs. The analytical uncertainty of each determination is set to be 25 $^{14}$C yrs. Effectively, this example is analogous to the atmospheric dataset `two_normals` but the determinations are instead sampled from the marine environment.   

We wish to investigate if we can reconstruct the underlying mixture of two normals calendar age distribution, from just the $^{14}$C values. For example, if we use the recommended Polya Urn approach:

```{r example_two_normals_code, eval = FALSE}
polya_urn_output <- PolyaUrnBivarDirichlet(
  rc_determinations = two_normals_marine$c14_age, 
  rc_sigmas = two_normals_marine$c14_sig, 
  calibration_curve = marine20, 
  delta_r = -50, 
  delta_r_sig = 30)

polya_DPMM_plot <- PlotPredictiveCalendarAgeDensity(
  output_data = polya_urn_output,
  show_SPD = TRUE)
```
```{r example_two_normals_marine, echo = FALSE, out.width="100%"}
oldpar <- par(no.readonly = TRUE)

# Fit the Polya urn mode to two_normals_marine 14C data
polya_urn_output <- PolyaUrnBivarDirichlet(
  rc_determinations = two_normals_marine$c14_age, 
  rc_sigmas = two_normals_marine$c14_sig, 
  calibration_curve = marine20, 
  delta_r = -50, 
  delta_r_sig = 30,
  show_progress = FALSE)

polya_DPMM_plot <- PlotPredictiveCalendarAgeDensity(
  output_data = polya_urn_output,
  show_SPD = TRUE)


predictive_density <- polya_DPMM_plot$predictive_density

par(new = TRUE, 
    mgp = c(3, 0.7, 0),
    xaxs = "i",
    yaxs = "i",
    mar = c(5, 4.5, 4, 2) + 0.1,
    las = 1)
xlim <- rev(range(predictive_density[[1]]$calendar_age_BP))
ylim <- c(0, 3 * max(predictive_density[[1]]$density_mean))
plot(
  NULL,
  NULL,
  type = "n",
  ylim = ylim,
  xlim = xlim,
  axes = FALSE,
  xlab = NA,
  ylab = NA,
  xaxs = "i",
  yaxs = "i")
# Show true underlying calendar age density
weights_true <- c(0.45, 0.55)
cluster_means_true_calBP <- c(3500, 5000)
cluster_precisions_true <- 1 / c(200, 100)^2

# Find and plot true exact density
truedens <- function(t, w, truemean, trueprec) {
  dens <- 0
  for(i in 1:length(w)) {
    dens <- dens + w[i] * dnorm(t, mean = truemean[i], sd = 1/sqrt(trueprec[i]))
  }
  dens
}
curve(truedens(
  x,
  w = weights_true,
  truemean = cluster_means_true_calBP,
  trueprec = cluster_precisions_true),
  from = 2500, to = 7000, n = 401,
  lwd = 2,
  col = "red", add = TRUE)

# Reset plotting parameters
par(oldpar)
```
*Note:* Here we have also shown the underlying calendar age distribution, consisting of a mixture of two normals, that was used for simulating the data in red. 


## Poisson process rate estimation from a set of marine ^14^C samples
If we have a set of marine ^14^C samples which share a common $\Delta R$ then we can use the `PPcalibrate()` function, specifying the value of $\Delta R$ and its corresponding uncertainty as additional arguments, and stating we wish to use the marine radiocarbon calibration curve. 

### Example - Uniform Phase Marine Data 
The dataset `pp_uniform_phase_marine` contains 40 simulated marine ^14^C samples with underlying calendar ages that have been sampled uniformly from the calendar period from [550, 500] cal yr BP. The corresponding $^{14}$C determinations are then simulated based upon the Marine20 calibration curve with an assumed $\Delta R_{20}$ of $100 \pm 20$ $^{14}$C yrs. The analytical uncertainty of each determination is set to be 15 $^{14}$C yrs. Effectively, this example is analogous to the amospheric dataset `pp_uniform_phase` but the determinations are instead sampled from the marine environment.   

We wish to investigate if we can reconstruct the underlying uniform phase calendar age distribution, from just the $^{14}$C values.

```{r example_uniform_marine, out.width="100%"}
# Fit the Poisson process model to marine 14C data
PP_fit_output_marine <- PPcalibrate(
    rc_determinations = pp_uniform_phase_marine$c14_age,
    rc_sigmas = pp_uniform_phase_marine$c14_sig,
    calibration_curve = marine20,
    delta_r = 100, 
    delta_r_sig = 20,
    n_iter = 1e5,
    show_progress = FALSE)

# Plot the posterior mean rate
posterior_mean_marine_plot <- PlotPosteriorMeanRate(PP_fit_output_marine)

# Add shading to show the period from which the underlying data were sampled
AddShadingPlot(posterior_mean_marine_plot,
    x_start = 550, x_end = 500,
    col = "red")

```
Here, on the radiocarbon age axis, we show the observed $^{14}$C marine determinations in black, and the adjusted determinations (after applying a $\Delta R = 100 \pm 20$ ^14^C yrs) are shown in green. Intuitively, we jointly calibrate and summarise the green (adjusted) values against the Marine20 curve.

In this example, we can see that the Poisson process summarisation still provides a posterior estimate for the sample occurrence rate that is consistent with the true (simulated) uniform phase [550, 500] cal yr BP calendar age distribution. The times of the changepoints are however less certain and can extend beyond this short 50 cal yr period (see also the plot of changepoint locations shown below). This is perhaps to be expected since the Marine20 curve is less precise and so a broader range of calendar ages are consistent with the simulated ^14^C data. 

We can also access the estimated posterior mean rate
```{r example_uniform_marine_post_mean, out.width="100%"}
# Access the actual posterior
posterior_marine_rate <- posterior_mean_marine_plot$posterior_rate
head(posterior_marine_rate)
```

<!-- around 500 cal yr BP when the creation of samples end is accurately estimated. The older changepoint at 550 cal yr BP has a less certain calendar age estimate. This is to be expected since we see there is a small plateau in the Marine20 calibration curve from ca. 600--550 cal yr BP which aligns with the $\Delta R$ adjusted $^{14}$C determinations.   -->

### Plotting the number of changepoints and their locations
The object `PP_fit_output_marine` can also be accessed directly or used with the other in-built plotting functions just as if the underpinning $^{14}$C data were atmospheric, for example, we can access the posterior estimate of the number of changepoints:

```{r example_uniform_marine_n_changes, out.width="100%"}
PlotNumberOfInternalChanges(PP_fit_output_marine)
```

and the posterior estimate of the location of those changepoints (conditional on their number):

```{r example_uniform_marine_changepoints, out.width="100%"}
posterior_changepoint_marine_plot <- PlotPosteriorChangePoints(PP_fit_output_marine)

# Add lines to show the times of the changepoints in
# the underlying distribution from which the data were 
# simulated at 500 and 550 cal yr BP
AddLinePlot(
     posterior_changepoint_marine_plot,
     v = 550,
     col = "purple",
     lwd = 2,
     lty = 2)

AddLinePlot(
     posterior_changepoint_marine_plot,
     v = 500,
     col = "purple",
     lwd = 2,
     lty = 2)

AddTextPlot(posterior_changepoint_marine_plot,
    x = 550, y = 0.0165,
    labels = expression(paste("550 cal yrs BP")),
    cex = 0.7,
    pos = 2,
    offset = 0.2,
    col = "purple")

AddTextPlot(posterior_changepoint_marine_plot,
    x = 500, y = 0.0165,
    labels = expression(paste("500 cal yrs BP")),
    cex = 0.7,
    pos = 4,
    offset = 0.2,
    col = "purple")

```
We can also plot the heights (i.e. rates) in each interval 
```{r example_uniform_marine_heights, out.width="100%"}
posterior_heights_marine_plot <- PlotPosteriorHeights(PP_fit_output_marine)
```


## References

