---
title: "Independent Single Sample Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Independent Single Sample Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(carbondate)
```

# Introduction

This package is concerned with the calibration of multiple related ^14^C determinations, it can also be useful to calibrate a single calibration independently, for example to estimate initial values for the sampler, or for comparison. To calibrate a single independent determination using the provided IntCal20 calibration curve [@reimer2020], run the following:

```{r single_calibration, out.width="100%", fig.width=10, fig.height=8}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE)
```   


## Implementing SH calibration

To change to using the Southern Hemisphere calibration curve SHCal20 [@hogg2020], we can modify the function arguments accordingly: 

```{r single_calibration_SH, out.width="100%", fig.width=10, fig.height=8}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = shcal20,
  plot_output = TRUE)
```
**Note:** This simulated SH calibration, of the same ^14^C determination as above, provides a slightly more recent estimate for the calendar due to the offset (the interhemispheric ^14^C gradient) between ^14^C levels in the Southern and Northern Hemisphere. 


## References


