---
title: "Independent Single Sample Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Independent Single Sample Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev='png',
  fig.width=7, 
  fig.height=5.5,
  dev.args=list(antialias = "none")
)
```

```{r setup}
library(carbondate)
```

# Introduction

This package is primarily concerned with the calibration, and summarisation, of multiple related ^14^C determinations. With such a set of samples, which are believed to arise from a common calendar distribution related to the culture/context under study, calibration should be performed jointly. This will allow calendar age information to be shared between the samples during the calibration process leading to improved overall calibration accuracy. It also allows us to combine and summarise the calendar age information in the entire set. 

However, our library also provides functions for calibrating a single ^14^C sample independently. This can be useful if one does only have a single sample, or to provide comparison with the joint calibration approaches. 

## Independent Calibration of Single ^14^C Samples

To calibrate a single independent determination using the provided IntCal20 calibration curve [@reimer2020], run the following:

```{r single_calibration, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE)
```   


### Implementing SH calibration

To change to using the Southern Hemisphere calibration curve SHCal20 [@hogg2020], we can modify the function arguments accordingly: 

```{r single_calibration_SH, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = shcal20,
  plot_output = TRUE)
```

**Note:** This simulated SH calibration, of the same ^14^C determination as above, provides a slightly more recent estimate for the calendar due to the offset (the interhemispheric ^14^C gradient) between atmospheric ^14^C levels in the Southern and Northern Hemisphere. 

## Selecting the calendar age scale when plotting

All the calibration curves, and the outputs from our functions, are provided on the cal yr BP scale (where 0 cal yr BP = 1950 cal AD). However, for the plots, the user can alter the calendar age scale shown through the `plot_cal_age_scale` variable. The default is to plot the calendar ages on the _cal yr BP_ scale (e.g., see above). To instead plot in _cal AD_, set `plot_cal_age_scale = "AD"`; while for _cal BC_ set `plot_cal_age_scale = "AD"`. For example, 

```{r single_calibration_AD, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE,
  plot_cal_age_scale = "AD")
``` 


## References


