---
title: "Independent Single Sample Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Independent Single Sample Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev='png',
  fig.width=7, 
  fig.height=5.5 # ,
  # dev.args=list(antialias = "none")
)
```

```{r setup}
library(carbondate)
```

# Introduction

This package is primarily concerned with the calibration, and summarisation, of multiple related ^14^C determinations. With such a set of samples, which are believed to arise from a common calendar distribution related to the culture/context under study, the calibration of those samples should be performed jointly. This will allow calendar age information to be shared between the samples during the calibration process leading to improved overall calibration accuracy. It also allows us to combine and summarise the calendar age information provided by the entire set. 

However, our library also provides functions for calibrating a single ^14^C sample independently. This can be useful if one does only have a single ^14^C sample, or to provide comparison with the joint calibration approaches. 

## Independent Calibration of Single ^14^C Samples

To calibrate a single independent determination using the provided IntCal20 calibration curve [@reimer2020], run the following:

```{r single_calibration, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE)
```   


### Implementing SH calibration

To change to using the Southern Hemisphere calibration curve SHCal20 [@hogg2020], we can modify the function arguments accordingly: 

```{r single_calibration_SH, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = shcal20,
  plot_output = TRUE)
```

**Note:** This simulated SH calibration, of the same ^14^C determination as above, provides a slightly more recent estimate for the calendar due to the offset (the interhemispheric ^14^C gradient) between atmospheric ^14^C levels in the Southern and Northern Hemisphere. 

### Implementing Marine calibration

To change to using the Marine20 calibration curve [@heaton2020], we can alter the argument to select the inbuilt `marine20` curve. To perform marine calibration, we must also provide an estimate for the localised $\Delta R$ (and its $1\sigma$-uncertainty) by specifying `delta_r` and `delta_r_sig`: 

```{r single_calibration_Marine, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = marine20,
  delta_r = 150,
  delta_r_sig = 50, 
  plot_output = TRUE)
```

Here the radiocarbon age axis shows the analytical ^14^C determination in red, while the adjusted determination (after applying a $\Delta R = 150 \pm 50$ ^14^C yrs) is shown in green. Intuitively, we calibrate the green (adjusted) value against the Marine20 curve. 

**Note:** One can also use the `delta_r` and `delta_r_sig` arguments to specify a general offset to a calibration curve (i.e., these arguments can also be used when using the IntCal and SHCal curves if the sample is thought to be offset).    


## Selecting the calendar age scale when plotting

All the calibration curves, and the outputs from our functions, are provided on the cal yr BP scale (where 0 cal yr BP = 1950 cal AD). However, for the plots, the user can alter the calendar age scale shown through the `plot_cal_age_scale` variable. The default is to plot the calendar ages on the _cal yr BP_ scale (e.g., see above). To instead plot in _cal AD_, set `plot_cal_age_scale = "AD"`; while for _cal BC_ set `plot_cal_age_scale = "BC"`. For example, 

```{r single_calibration_AD, out.width="100%"}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE,
  plot_cal_age_scale = "AD")
``` 

## Annotating the plots with text, lines and shading
We have created three functions:

* `AddLinePlot()` - adds lines (vertical or horizontal)
* `AddTextPlot()` - adds text
* `AddShadingPlot()` - adds shading

that allow you to annotate the posterior calendar age density plot. For example:

```{r annotate_posterior_plot, out.width="100%"}
# Create initial plot
calibration_result_plot <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20,
  plot_output = TRUE)

# Add solid vertical line at 1370 cal yr BP
AddLinePlot(calibration_result_plot,
  v = 1370,
  col = "darkgreen",
  lwd = 2,
  lty = 1)

AddTextPlot(calibration_result_plot,
  x = 1370, y = 1300,
  labels = expression(paste("1370 cal yrs BP")),
  cex = 0.7,
  pos = 4, # Places the text to the right
  offset = 0.2,
  col = "darkgreen")

# Add a dot-dashed horizontal line at 1220 14C yrs BP
AddLinePlot(calibration_result_plot,
  h = 1220,
  col = "darkorange",
  lwd = 2,
  lty = 4)
    
AddTextPlot(calibration_result_plot,
  x = 1250, y = 1220,
  labels = expression(paste("1220", " "^14, "C ", "yrs BP")),
  cex = 0.9,
  pos = 1, # Places text below 
  offset = 0.2,
  col = "darkorange")

# Add light gray grid lines to entire plot
AddLinePlot(calibration_result_plot,
  h = seq(1000, 1700, by = 50),
  v = seq(1100, 1600, by = 50),
  col = "lightgray",
  lty = 3)

# Add a shaded box to cover a period
AddShadingPlot(calibration_result_plot,
  x_start = 1250, x_end = 1200,
  y_start = 1300, y_end = 1400,
  col = "green", alpha = 0.2)

```



## References


