---
title: "Introduction to carbondate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to carbondate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(carbondate)
set.seed(5)
```

# Introduction

The **carbondate** package contains functions for analysing sets of related radiocarbon (^14^C) determinations. Due to fluctuations in past radiocarbon levels, all ^14^C determinations need to be calibrated (converted) in order to be understood on the calendar age scale. In many cases, we wish to calibrate a set of samples which are known to be related to one another (for example, arising from a particular site, or set of sites, populated by a particular culture). Each sample has an unknown calendar age but, since they are related, these calendar ages are assumed to arise from the same unknown shared calendar age density.

This library serves two purposes:

- calibration of the set of ^14^C determinations can be improved significantly by incorporating the knowledge that the samples are related;
- summarisation of the underlying calendar age distribution from which the samples arise provides a potentially useful proxy for activity, population size, the frequency of past events. This approach uses the principle of _dates-as-data_ [@rick1987]

We provide two different, albeit ideologically linked, approaches to achieve this:

- A non-parametric Bayesian density estimation approach using a Dirichlet process mixture model. This is a statistically-rigorous alternative to summed probability distributions (SPDs) and kernel density estimates (KDE model).
- A inhomogeneous Poisson process/Changepoint approach using Reversible Jump Markov Chain Monte Carlo (RJ-MCMC) to estimate the changing rate at which the ^14^C samples occur over calendar time.    
Separate vignettes are provided to describe each approach.

## Dates-as-Data -- Considering frequency of samples as a proxy for activity 

A commonly-used approach to estimate changes in the frequency of past events or the size of populations looks at variations in the rate of archaeological and environmental samples (e.g., charcoal from fires, human/animal bones, or other evidence of occupation) found at a site over time. Time periods with large numbers of samples suggest increased activity, while those with few samples indicate a reduced level of activity. This paradigm is known as _dates-as-data_ [@rick1987]. 

The reliability of such a _dates-as-data_ approach is highly dependent upon our ability to estimate the calendar ages of the discoveries. Most archaeological/environmental dates are obtained using radiocarbon . The need for calibration of these ^14^C samples introduces considerable uncertainties in the resultant calendar ages and complicates the identification of changepoints in the calendar year rates at which samples occur.

In this library, we provide two theoretically-underpinned approaches to overcome these challenges, each of which has its own set of vignettes:

- A statistically-rigorous alternative to summed probability distributions (SPDs) that models the samples as arising from a mixture distribution containing an unknown number of distinct clusters. [Non-parametric summed density estimation   vignette](Non-parametric-summed-density.html) ;   
- A changepoint-based approach that models the occurrence of the ^14^C samples as an inhomogeneous Poisson process. We aim to reconstruct how the rate at which the samples occur varies over calendar time, and to identify if there are statistically significant changepoints (i.e., specific calendar times at which the rate of samples changes abruptly).  [Poisson process modelling vignette](Poisson-process-modelling.html) 


## Additional Information

The library also provides some joint data (which can be usd by either method) as described below 


## Calibration Curve

Given a set of reference objects for which we have both ^14^C measurements and independently known (or estimated) calendar ages, we can create what is known as a calibration curve. This calibration curve is a mapping providing, for an object of true calendar age $\theta$, the corresponding radiocarbon age $\mu(\theta)$. The radiocarbon determination of any individual sample is assumed to be a noisy observation of this, i.e., 
$$
X_i \sim N(\mu(\theta), \sigma_{\textrm{lab}}^2)
$$

Given a undated object for which we obtain a ^14^C determination one can estimate its calendar age by inverting this mapping. A calibration curve is a required input for all the functions in this package.

The internationally-ratified, standard for the radiocarbon calibration curve is  known as IntCal, for which regular updates are provided by the IntCal working group. IntCal20 [@reimer2020] is the current version agreed for use by the community. This provides pointwise estimates of the mean $m(\theta)$ and sd $\rho(\theta)$ of the calibration curve, which can be integrated out during calibration, i.e.,:
\begin{align*}
\mu(\theta) | \theta & \sim N(m(\theta), \rho(\theta)^2), \textrm{ and} \\
\Rightarrow X_i & \sim N(m(\theta), \rho(\theta)^2 + \sigma_{\textrm{lab}}^2)
\end{align*}

The curve data `intcal20` is provided in this package. However it is possible to load and use another calibration curve if you wish.

## Example Data

There are some example ^14^C data sets provided in the package, each of which contain a number of radiocarbon determinations and uncertainties from real-life cases:

-   `kerr`: 255 radiocarbon determinations collated by Kerr and McCormick related to the building and use of raths in Ireland [@kerr2014].

-   `armit`: 2021 radiocarbon determinations collated by Armit et al. from archaeological groups operating in Ireland, to investigate whether a wetter environment around 2700 cal yr BP led to a population collapse [@armit2014].

-   `buchanan`: 628 radiocarbon determinations collated by Buchanan et al. representing the ages of distinct archaeological sites found across Canada and North America during the time of the palaeoindians [@buchanan2008].

-   `alces`, `bison`, `cervus`, `equus`, `human`, `mammuthus`: radiocarbon determinations related to a variety of megafauna (and humans) from Yukon and Alaska [@guthrie2006]. Samples are restricted to those between 25,000--6000 ^14^C yrs BP. 

# Independent calibration

Although this package is concerned with the calibration of multiple related ^14^C determinations it can also be useful to calibrate a single calibration independently, for example for estimating initial values for the sampler, or for comparison. To calibrate a single independent determination (using the provided IntCal20 calibration curve), run the following:

```{r single_calibration, out.width="100%", fig.width=10, fig.height=8}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20)
plot(
  calibration_result,
  type="l", 
  xlab = "Calendar age (yr BP)", 
  ylab = "Probability", 
  xlim=c(1600, 1200),
  yaxt="n")
```

For all functions the radiocarbon determinations can be entered as F^14^C concentrations or as the ^14^C age BP. The flag `F14C_inputs` used above (and in other functions) lets the function know what units have been used for the radiocarbon determinations.

# Summed Probability Distributions

The most common approach to summarise calendar age information of multiple ^14^C determinations is via summed probability distributions (SPD). These are **not statistically valid** estimators of the calendar age of a potential future sample. They should not be used.  


Here, the posterior calendar age density of each object is first calculated independently) from the others as using the function above. These individual densities are then summed/averaged to give an SPD estimate. 

SPDs are not statistically valid estimators of the calendar age of a potential future object. The independence assumed in the separate calibration of each sample, followed by subsequent summarisation, generates a contradiction. The function is provided here as a useful comparison with the other routines.

To calculate the SPD for a set of radiocarbon determinations (here we use the example dataset `armit`) see the example below, where we also plot the results.

```{r find_spd, out.width="100%", fig.width=10, fig.height=8}
spd <- FindSummedProbabilityDistribution(
  calendar_age_range_BP = c(1000, 4500), 
  rc_determinations = armit$c14_age, 
  rc_sigmas = armit$c14_sig, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20)
plot(
  spd,
  type="l", 
  xlab = "Calendar age (yr BP)", 
  ylab = "Probability", 
  xlim=c(4500, 1000),
  yaxt="n")
```

Note that the summary functions for plotting the predictive joint calendar age density [described below](#density-estimate-to-summarise-objects) can also optionally plot the SPD without having to calculate it separately first.



# References
