---
title: "Poisson Process Modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson Process Modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(carbondate)
set.seed(15)
```

# Introduction

<!-- badges: start -->
<!-- badges: end -->

Radiocarbon Calibration Of Multiple $^{14}$C Samples Using Reversible Jump MCMC and a Poisson process

The goal of RJradiocarbon is to model the occurrence of radiocarbon samples as a (variable rate) Poisson process. Given a set of multiple $^{14}$C samples, 
$$X_1, \ldots, X_n$$
the function `PPcalibrate` will simultaneously:

- calibrate them to provide posterior calendar ages estimates $\theta_1, \ldots, \theta_n$ 
- estimate the rate $\lambda(t)$ of the underlying process. This estimate $\hat{\lambda}(t)$ may provide a useful proxy to make archaeological and/or environmental inference, e.g., to model changes in culture, or useage, or population size over time. 


The (calendar) rate at which the samples occur is modelled as a Poisson process with a variable rate $\lambda(t)$. Specifically, $\lambda(t)$ is modelled as a piecewise constant function with an unknown number of changepoints:
$$
\lambda(t) = \left\{ \begin{array}{ll} 
  h_i \quad & \textrm{if } s_i \leq t < s_{i+1} \\
  0 & \textrm{otherwise}
  \end{array} \right.
$$
where $T_A < s_1 < s_2 < \ldots < s_k < T_B$ are the calendar ages at which there are changes in the Poisson process rate, and $T_A$ and $T_B$ are bounding calendar ages. We use reversible jump MCMC (RJMCMC) to alternate between calibrating the samples and updating the estimate of the underlying rate lambda. 


## Example

This is a basic example which shows you how to solve a common problem. In this simulated example, we will create 40 artificial samples with $^{14}$C determinations corresponding to the calendar period from [550, 500] cal yr BP. Effectively, the underlying model is a Poisson process with a rate 
$$
\lambda(t)  \left\{ \begin{array}{ll}
  h_1 = 0 & \textrm{if } 400 \leq t < 500 \textrm{ cal yr BP} \\
  h_2 \qquad & \textrm{if } 500 \leq t < 550 \textrm{ cal yr BP}\\
  h_3 = 0 & \textrm{if } 550 \leq t < 700 \textrm{ cal yr BP} 
  \end{array} \right.
$$
We will then investigate if we can reconstruct this underlying information, from just the $^{14}$C values, if we only knew that the calendar ages of the samples lay between the bounding [700,400] cal yr BP. 

```{r example, out.width="100%", fig.width=10, fig.height=8}
# Set initial values 
n_observed <- 40
n_iter <- 100000
n_thin <- 10
rc_sigmas <- rep(15, n_observed)
calendar_grid_resolution <- 1

# Create artificial rc_determinations
calendar_age_range <- c(400, 700) 
observed_age_range <- c(500, 550) 
true_theta <- seq(from = observed_age_range[1],
                  to = observed_age_range[2],
                  length = n_observed)
intcal_mean <- approx(x = intcal20$calendar_age_BP,
                      y = intcal20$c14_age,
                      xout = true_theta)$y
intcal_sd <- approx(x = intcal20$calendar_age_BP,
                    y = intcal20$c14_sig,
                    xout = true_theta)$y
rc_determinations <- rnorm(n = n_observed,
                                  mean = intcal_mean,
                                  sd = sqrt(rc_sigmas^2 + intcal_sd^2))

# Fit the model
PP_fit_output <- PPcalibrate(
    rc_determinations = rc_determinations,
    rc_sigmas = rc_sigmas,
    calibration_curve = intcal20,
    calendar_age_range = calendar_age_range,
    calendar_grid_resolution = calendar_grid_resolution,
    n_iter = n_iter,
    n_thin = n_thin,
    show_progress = FALSE)
```

This creates an object `PP_fit_output` which you can access directly or use the in-built plotting functions.

### Plotting the posterior mean estimate of $\lambda(t)$  
To plot the posterior estimate of $\lambda(t)$ you should use the in-built function: 
```{r meanrate, out.width="100%", fig.width=10, fig.height=8}
PlotPosteriorMeanRate(PP_fit_output)
```

This plots the posterior mean rate of sample occurrence (shown in purple). The $^{14}$C determinations ($X_1, \ldots, X_n$) are shown on the y-axis as a rug. The rug on the x-axis plots the posterior mean of the corresponding calendar ages (i.e., $\theta_1, \ldots, \theta_n$) after calibration for illustration purposes. **Note**: In some cases (in particular when an individual calibrated age estimates $\theta_i$ is bimodal) the posterior calendar mean will not be a good summary.  

As you can see, the posterior estimate of $\lambda(t)$ identifies the calendar period from which the hypothetical samples have been created [550, 500] cal yr BP. Here, $\hat{\lambda}(t)$ is non-zero, while for the other time periods.

### Plotting the estimated number of changepoint
You can plot the posterior estimate for the number of internal changes in the rate $\lambda(t)$ using the command:  
```{r changepoint number, out.width="100%", fig.width=10, fig.height=8}
PlotNumberOfInternalChanges(PP_fit_output)
```

Here, we can see that the method estimates that there are most likely to be two changes. This is clearly an overestimate of the number of changes --- in our underlying data there are only two changes, a step up from a occurrence rate of zero samples/cal yr to a rate of *ca.* 0.8 samples/cal yr at 550 cal yr BP, and then a step back down to return to a sample occurrence rate of zero at 500 cal yr BP. 

This overestimation is largely a consequence of the prior we have chosen for the number of internal changes in the rate $\lambda(t)$. This has a default of 10 in `PPcalibrate()` to enable more flexibility in rate estimation when the true rate is not piecewise constant. The number of estimated changes will be pulled towards this value. 

### Plotting the estimates of changepoint locations
You can plot the locations of the changepoints, conditional on the number of changes currently estimated in the model. The default shows the locations when the model estimates $n_{\textrm{internal changes}} = 1, 2,$ and 3.    
```{r changepoint locations, out.width="100%", fig.width=10, fig.height=8}
PlotPosteriorChangePoints(PP_fit_output)
```

This figure can be interpreted as follows. For example, if the model thinks there are only two changepoints in the underlying occurrence rate of the samples, then these changepoints should be located around 500 cal yr BP and 550 cal yr BP (the two distinct dashed density estimates shown in dark green). If the model thinks there are three changepoints in the rate, then these should be located according to the location shown by the dotted red densities (note that here these three locations overlay one another, effectively adding a double changepoint at either 500 or 550 cal yr BP). Here, the model never thinks there should be only one changepoint (as indicated by the absence of a solid blue density, and the warning message).      

### Plotting the rate estimates conditional on the number of changes
Finally, you can plot the estimates of the rate of sample occurrence conditional on the number of changes estimate 
```{r changepoint rates, out.width="100%", fig.width=10, fig.height=8}
PlotPosteriorHeights(PP_fit_output)
```

Here, conditional on there being two changes in $\lambda(t)$ between 700--400 cal yr BP (so three different values for the rates) we have two rates of *ca.*  0 events per cal yr (these correspond to the periods 700--550 and 500--400 cal yr BP), and a period with rate *ca.* 0.6-0.8 events per cal yr (corresponding to the period from 550--500 cal yr BP)   
