---
title: "Temporary-Text"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporary-Text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(carbondate)
```

### TODO - Remove this at end


For all functions, the radiocarbon determinations can be entered as F^14^C concentrations or as the ^14^C age BP. The flag `F14C_inputs` used above (and in other functions) lets the function know what units have been used for the radiocarbon determinations.

## Implementing SH calibration

To change to using the Southern Hemisphere calibration curve SHCal20, we can modify the function arguments accordingly: 

```{r single_calibration_SH, out.width="100%", fig.width=10, fig.height=8}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = shcal20)
plot(
  calibration_result,
  type="l", 
  xlab = "Calendar age (cal yr BP)", 
  ylab = "Posterior Probability", 
  main = expression(paste("Calibration of ", 1413, " \u00B1 ", 25, " "^14, "C yr BP using SHCal20")), 
  xlim=c(1500, 1100))
```
**Note:** This simulated SH calibration, of the same ^14^C determination as above, provides a slightly more recent estimate for the calendar due to the offset (the interhemispheric ^14^C gradient) between ^14^C levels in the Southern and Northern Hemisphere. 



### Old Text

We model our unknown calendar age density $f(\theta)$  as an infinite mixture of individual calendar age clusters/phases. These individual calendar age clusters are normal densities that can have different locations and spreads. Such a model allows considerable flexibility into the estimation of the joint calendar age density $f(\theta)$, not only simple mixtures but also

In some cases, this mix of normal densities may represent true and distinct underlying normal archaeological phases, in which case additional practical inference may be possible. However this is not required for the method to provide good estimation, and the clusters will be used


### Intuitive Illustration
We model the underlying shared calendar age distribution as an unknown mixture of calendar age _clusters/phases_:
$$
f(\theta) = w_1 \textrm{Cluster}_1 + w_2 \textrm{Cluster}_2 + w_3 \textrm{Cluster}_3 + \ldots 
$$
where each _cluster_ has a normal distribution (with unknown mean $\mu_j$ and precision $\tau_j^2$).



Each object is then considered to be drawn from one of the (infinite) clusters which constitute the overall $f(\theta)$. The probability that it is drawn from a particular cluster will depend upon the relative weight given to that specific cluster. 

It will be more likely that an object will come from some clusters than others. Given an object belongs to a particular cluster, its prior calendar age will be normally distributed with the mean and variance of that cluster. The mean and variance of each individual normal cluster that constitutes the overall $f(\theta)$, together with the weightings associated to each cluster, will be estimated based upon the set of ^14^C determinations we observe.

For full technical details of the models used, and explanation of the model parameters, see [@heaton2021].


## Text from old Intro

# Additional Functions

## Independent calibration

Although this package is concerned with the calibration of multiple related ^14^C determinations, it can also be useful to calibrate a single calibration independently, for example to estimate initial values for the sampler, or for comparison. To calibrate a single independent determination (using the provided IntCal20 calibration curve), run the following:

```{r single_calibration, out.width="100%", fig.width=10, fig.height=8}
calibration_result <- CalibrateSingleDetermination(
  rc_determination = 1413, 
  rc_sigma = 25, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20)
plot(
  calibration_result,
  type="l", 
  xlab = "Calendar age (cal yr BP)", 
  ylab = "Posterior Probability",
  main = expression(paste("Calibration of ", 1413, " \u00B1 ", 25, " "^14, "C yr BP using IntCal20")), 
  xlim=c(1500, 1100))
  
```   

## Summed Probability Distributions
Currently, the most commonly-used approach to summarise calendar age information from multiple ^14^C determinations is via summed probability distributions (SPD). These are **not statistically valid** estimators of the calendar age of a potential future sample. It is our view that they should not be used in any dates-as-data approach to provide a population proxy.  

For an SPD, the posterior calendar age density of each object is first calculated independently) from the others as using the function above. These individual densities are then summed/averaged to give an SPD estimate. The independence assumed in the separate calibration of each sample, followed by subsequent summarisation, generates a contradiction. 

Additionally, the SPDs approach fundamentally does not model the samples in the calendar age domain. Consequently, it is also not able to deal with inversions in the calibration curve where there are multiple disjoint calendar periods which are consistent with the observed determinations; or with plateau periods. 

The SPD function is **ONLY** provided here as a comparison with the other routines. To calculate the SPD for a set of radiocarbon determinations (here we use the example dataset `armit`) see the example below, where we also plot the results.

```{r find_spd, out.width="100%", fig.width=10, fig.height=8}
spd <- FindSummedProbabilityDistribution(
  calendar_age_range_BP = c(1000, 4500), 
  rc_determinations = armit$c14_age, 
  rc_sigmas = armit$c14_sig, 
  F14C_inputs = FALSE, 
  calibration_curve = intcal20)
plot(
  spd,
  type="l", 
  xlab = "Calendar age (cal yr BP)", 
  ylab = "SPD Probability", 
  xlim=c(4500, 1000),
  yaxt="n")
```

**Note:** The summary functions for plotting the predictive joint calendar age density using the rigorous Bayesian non-parametric alternative to SPDs (described in the vignette [Non-parametric Summed Density](Non-parametric-summed-density.html)) can also optionally plot the SPD without having to calculate it separately first.

### Website stuff

Due to fluctuations in past radiocarbon levels, all 14C determinations need to be calibrated (converted) in order to be understood on the calendar age scale. In many cases, we wish to calibrate a set of samples which are known to be related to one another (for example, arising from a particular site, or set of sites, populated by a particular culture). Each sample has an unknown calendar age but, since they are related, these calendar ages are assumed to arise from the same unknown shared calendar age density.

### Text from independent calibration

Introduction 

This package is concerned with the calibration of multiple related 14C determinations. However it can also be useful to calibrate a single calibration independently, for example to estimate initial values for the sampler, or for comparison. 

### Text from determining convergence

summary 

model the posterior calendar ages are not suitable parameters to
use - rather we are interested in the number and distribution of clusters. However, they cannot be easily
compared using this diagnostic as the number and identity of clusters can change with each iteration. For this model,
the predictive density is a much more useful indicator.

## References
